Contracts for unions and intersections are hard for Nickel

### 2021-06-07 paper.tex
Reworking of section 4 (last subsection on semantics, WIP)

A natural - but naive - implementation of intersection contracts could be the
following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
and \nickel{B} sequentially:

\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Naive
decomposition}]
M | A @& B (*@ $\simeq$ @*) (M | A) | B
\end{lstlisting}

This intuition works for simple contracts: checking that \nickel{x | Natural @&
Odd} amounts to check that \nickel{x | Natural} and \nickel{x | Odd}.
Unfortunately, this doesn't scale to higher-order contracts.
Figure~\ref{fig:inter-contracts} illustrates the use of an intersection to model
a simple overloading of the identity function. If we were to apply the naive
equivalence above, the argument \nickel{1} would fail the \nickel{Str -> Str}
contract. The naive decomposition actually implies

\begin{lstlisting}[language=nickel,frame=none,numbers=none]
(A -> B) @& (C -> D) (*@ $\simeq$ @*) (A @& C) -> (B @& D)
\end{lstlisting}

This isn't the correct semantics for overloading. With this implementation, our
contract for identity from Figure~\ref{fig:inter-contracts} would always
fail because no argument satisfies \nickel{Num @& Str}.

@@

\begin{figure}[h]
% \begin{lstlisting}[language=nickel, title=Implementation]
% x | A @& B (*@ $\simeq$ @*) (x | A) | B
% \end{lstlisting}
\begin{lstlisting}[language=nickel]
let g | Num -> Num @& Str -> Str
      = fun x => x in
g 1
\end{lstlisting}
\caption{In}
\label{fig:inter-contracts}
\end{figure}

@@

@@

\subsection*{Union and intersection are not a silver bullet}

Before wrapping up this section, we would like to show that maybe, union and intersection
contracts are not exactly what we would want them to be.
As stated throughout this work, programmers coming from dynamic languages
are accustomed to perform safety checks (data validation) with
great amounts of flexibility and the whole power of their
language of choice at their disposal.

For instance, a Python programmer may have no problem in checking
at runtime, that a given variable is an integer if a given flag is
not set, as shown in Figure \ref{fig:python-assertion};
while a Haskell programmer might close her eyes and leave the room
instantly.

\begin{figure}[h]
\begin{lstlisting}[language=python]
result = 42 if flag else "Disabled"
...
assert (
  (not flag)
  or isinstance(result, int)
)
\end{lstlisting}
\caption{Asserting at runtime the type of a variable}
\label{fig:python-assertion}
\end{figure}

These expectations are not really met by how unions and intersections
contracts are supposed to work.
For example, consider the Nickel code shown on Figure
\ref{fig:intersection-distrbution}, at first sight, this code should
complete execution successfully, returning \nickel{(1, 1)},
however, under both the Keil and Thiemann, as well as the
Williams, Morris, and Wadler works, this program will fail, with a violation
of the contract imposed to \nickel{g}.

@@

The reason this happens is that there is a fundamental property of how
intersection contracts behave: the choice of which branch to choose is made
exactly when the intersection is opened up, and not delayed as much as possible,
as one would like.  This means, that the contract presented on Figure
\ref{fig:intersection-distrbution}, \nickel{Num -> Num -> Num @& Num
-> Bool -> Num}, will not behave the same way that \nickel{Num ->
(Num -> Num @& Bool -> Num)}, when, with an intuitive view of how
intersections should behave, these two expressions should be logically
equivalent.

It is worth noting, that intersection types do have this property,
since subtyping would allow to correctly type check a term like
\nickel{g 1} as both \nickel{Num -> Num} as well as
\nickel{Bool -> Num}, and then deducing that \nickel{g 1}
can be given the type \nickel{Num -> Num @& Bool -> Num}.

>>>>

A natural - but naive - implementation of intersection contracts could be the
following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
and \nickel{B} sequentially, resulting in the naive decomposition rule of
Figure~\ref{fig:naiv-impl}.

\begin{figure}[h]
\label{fig:naive-impl}
\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Naive
decomposition}]
M | A @& B (*@ $\simeq$ @*) (M | A) | B
\end{lstlisting}
\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Exchange law}]
(A -> B) @& (C -> D) (*@ $\simeq$ @*) (A @& C) -> (B @& D)
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Overloaded identity}]
let g | Num -> Num @& Str -> Str
      = fun x => x in
g 1
\end{lstlisting}
\caption{Naive implementation of intersection}
\end{figure}

This intuition works for simple contracts: checking that \nickel{x | Natural @&
Odd} amounts to check that \nickel{x | Natural} and \nickel{x | Odd}.
Unfortunately, this doesn't scale to higher-order contracts.  The overloaded
identity example of Figure~\ref{fig:naive-impl} illustrates the use of an
intersection to model a simple overloading of the identity function. If we were
to apply the naive decomposition, the argument \nickel{1} would fail the
\nickel{Str -> Str} contract and abort the execution. Perhaps the exchange rule
given in Figure~\ref{fig:naiv-impl}, which is a direct consequence of the naive
decomposition, illustrates the issue better. It is clear that this exchange law
isn't the right semantics for overloading. With this law, the contract for
overloaded identity of Figure~\ref{fig:naiv-impl} would always fail because
no argument can satisfy \nickel{Num @& Str}.

@@

@@

\subsection*{Challenging the semantics}

@@

Previous points of this section were about how implementing the expected
semantics for union and intersection has a profound impact on the language. Let
us end the section by investigating a different aspect: does the
semantics correspond to the programmer's intuition?

But which semantics? We deemed the exchange law of Figure~\ref{fig:naive-impl}
as clearly wrong with respect to overloading. There is nothing wrong about it
per se : we could chose a semantics for intersections that validate this rule.
However, the resulting system wouldn't capture overloading, and is arguably not
very interesting nor useful. In~\ref{KeilThiemannUnionIntersection}, the authors
argue for basing the semantics of union and intersection contracts on union and
intersection type systems, where they have a simple and natural definition from
which the other properties follow. This approach makes sense, and has been
continued in future works, and in particular in~\ref{RootCauseOfBlame}.

But even this semantics is no panacea. Figure~\ref{fig:intersection-distrbution}
shows an example that intuitively looks like it should succeeed, returning
\nickel{(1, 1)}. However, under the semantics used both
in~\ref{KeilThiemannUnionIntersection} and~\ref{RootCauseOfBlame}, this program
fails with a contract violation.

% Before wrapping up the section, we would like to show that maybe, union and
% intersection contracts are not exactly what we would want them to be.  As stated
% throughout this work, programmers coming from dynamic languages are accustomed
% to perform safety checks (data validation) with great amounts of flexibility and
% the whole power of their language of choice at their disposal.
%
% For instance, a Python programmer may have no problem in checking at runtime,
% that a given variable is an integer if a given flag is not set, as shown in
% Figure \ref{fig:python-assertion}; while a Haskell programmer might close her
% eyes and leave the room instantly.
%
% \begin{figure}[h]
% \begin{lstlisting}[language=python]
% result = 42 if flag else "Disabled"
% ...
% assert (
%   (not flag)
%   or isinstance(result, int)
% )
% \end{lstlisting}
% \caption{Asserting at runtime the type of a variable}
% \label{fig:python-assertion}
% \end{figure}
%
% These expectations are not really met by how unions and intersections
% contracts are supposed to work.
% For example, consider the Nickel code shown on Figure
% \ref{fig:intersection-distrbution}, at first sight, this code should
% complete execution successfully, returning \nickel{(1, 1)},
% however, under both the Keil and Thiemann, as well as the
% Williams, Morris, and Wadler works, this program will fail, with a violation
% of the contract imposed to \nickel{g}.

@@

This discrepancy is inherent to the chosen semantics for intersections.  When
applying a function with an intersection contract, only one fixed component of
the intersection can be selected for the function.
one components of the intersection. Not that it must necessarily known right at the
application, but the result can't be used
functions, the choice of the component is fixed once for all when the function
is applied. In Figure~\ref{fig:intersection-distrbution}, the contract for
\nickel{h} can be either \nickel{Num -> Num} or \nickel{Bool -> Num}, but can't
be both. This is a manifestation of the following more general fact: the
distributivity law given in Figure~\ref{fig:distributivity-law} is not valid in
the semantics, while it is both intuitive and sound.\unsure{What means sound, in
this context?}

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
A -> B @& A -> C (*@ $\simeq$ @*) A -> (B @& C)
\end{lstlisting}
\caption{Distributivity law}
\label{fig:distributivity-law}
\end{figure}

%
% It is worth noting, that intersection types do have this property,
% since subtyping would allow to correctly type check a term like
% \nickel{g 1} as both \nickel{Num -> Num} as well as
% \nickel{Bool -> Num}, and then deducing that \nickel{g 1}
% can be given the type \nickel{Num -> Num @& Bool -> Num}.
%
